---
title: "Homework 2 - Data Wrangling 1"
author: "Sydney Ng (uni: sn2863)"
date: "due 09/30/2020 by 10:00 pm EDT"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

# Problem 1
```{r message = FALSE}
trashwheel_df <-
  read_excel("./hw2_prob1/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
             sheet = "Mr. Trash Wheel", 
             range = cell_cols("A:N")) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>% # specify a column to look at
  mutate(sports_balls = round(sports_balls),
         sports_balls = as.integer(sports_balls))
```

```{r}
precip_2018_df <-
  read_excel("./hw2_prob1/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
             sheet = "2018 Precipitation",
             skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2018) %>%
  relocate(year)

precip_2017_df <-
  read_excel("./hw2_prob1/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
             sheet = "2017 Precipitation",
             skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2017) %>%
  relocate(year)
```

Combining annual precipitation for 2017 and 2018.

```{r}
precip_df <- bind_rows(precip_2017_df, precip_2018_df)
  #mutate(month = lubridate::month(month))
         # month_name = month.name[month]) 
         # don't use square brackets Jeff doesn't like them

month_df <- tibble(month= 1:12, month_name = month.name)

left_join(precip_df, month_df, by = "month")
```

This data set contains information from the Mr. Trash Wheel collector in Baltimore, Maryland. As trash enters the inner harbor, it collects the trash and stores it in the dumpster.

Contains information on year, month, and the total trash collected. There are a total of **`r nrow(trashwheel_df)`** observations in our final Mr. Trash Wheel data set.

Additional data include month precipitation data for 2017 and 2018.

# Problem 2
```{r message = FALSE}
subway_df <-
  read_csv("./hw2_prob2/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line, station_name, station_latitude, station_longitude, route1:route11,
         entry, vending, entrance_type, ada) %>%
  mutate(entry = ifelse(entry == "YES", 1, 0))
  

head(subway_df) # just to check it out -- this was not asked for in hw
```

This data set contains information about the New York City subway system's station entrances and exits, as well as train routes. It also contains information about vending, ADA compliance, staffing, and exact geographical coordinates.

There are a total of `r nrow(subway_df)` observations in our final data set. We have specifically pulled line, station name, station latitude/longitude, routes, entry, vending, entrance type, and ADA compliance variables to work with.

* There are **`r nrow(distinct(subway_df, station_name, line))`** distinct subway stations by station name and line.
* There are **`r nrow(filter(subway_df, ada == TRUE))`** stations that are ADA compliant.
* The proportion of station entrances / exits without vending that allow entrance is **`r nrow(filter(subway_df, vending == "NO" & entry == 1))/nrow(filter(subway_df, vending == "NO"))`**.

#### Reformat the data so route number and route name are distinct variables

```{r}
# need to make all route# variable types consistent first before pivot_longer()
# decided to make all routes character variables, and dropped any missing routes

subway_reformat_df <- 
  subway_df %>%
  mutate(route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)) %>%
  pivot_longer(route1:route11, names_prefix = "route",
               names_to = "route_num", values_to = "route") %>%
  drop_na(route)

head(subway_reformat_df)
```

Further filtering by the A route and then answering questions:

```{r}
distinct_stations_A <-
  subway_reformat_df %>%
  filter(route == "A") %>%
  distinct(line, station_name, .keep_all = TRUE)

ada_stations_A <-
  subway_reformat_df %>%
  filter(ada == TRUE) %>%
  distinct(line, station_name, .keep_all = TRUE)
```

* There are **`r nrow(distinct_stations_A)`** distinct stations that serve the A train.
* Of the stations that serve the A train, **`r nrow(ada_stations_A)`** are ADA compliant.

# Problem 3
#### Clean the data in `pols-month.csv`
```{r message = FALSE}
pols_month_df <- 
  read_csv("./hw2_prob3/pols-month.csv") %>%
  separate(mon, c("year", "month", "day"), convert = TRUE)

month_df <- tibble(month= 01:12, month_name = month.name) # month names

pols_month_df <- 
  left_join(month_df, pols_month_df, by = "month") %>%
  mutate(president = case_when(prez_gop == 1 ~ "Republican",
                               prez_gop == 0 ~ "Democrat",
                               prez_gop == 2 ~ "Other")) %>%
  select(-prez_gop, -prez_dem, -day, -month) # removing variables
```

#### Clean the data in `snp.csv` using a similar process to the above
```{r message = FALSE}
snp_df <- 
  read_csv("./hw2_prob3/snp.csv") %>%
  separate(date, c("month", "day", "year"), convert = TRUE) %>%
  left_join(month_df, snp_df, by = "month") %>%
  select(year, month_name, close)
```

#### Tidy the unemployment data so that it can be merged with the previous datasets.
```{r message = FALSE}
unemployment_df <- 
  read_csv("./hw2_prob3/unemployment.csv") %>%
  pivot_longer(2:13, names_to = "month", values_to = "unemployment")

month_abb_df <- tibble(month = month.abb, month_name = month.name) # month names

unemployment_df <- 
  left_join(month_abb_df, unemployment_df, by = "month") %>%
  mutate(year = Year) %>% # renaming Year to year to stay consistent
  select(year, month_name, unemployment)
```

#### Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.
```{r}
FiveThirtyEight_df <- 
  left_join(snp_df, pols_month_df, by = c("year", "month_name")) %>%
  left_join(unemployment_df, by = c("year", "month_name"))

head(FiveThirtyEight_df)
```

The `pols-month` data set gives us information about the number of national politicians who are Democratic or Republican at any given date. Variables describe the number of representatives at different levels including the Executive level, senators and House Reps, and governors. The range of years is from `r min(pull(pols_month_df, year))` to `r max(pull(pols_month_df, year))`.

The `snp` data set gives us information about the S&P stock market index with a date variable and the closing value of the index on that date. The range of years is from `r min(pull(snp_df, year))` to `r max(pull(snp_df, year))`.

The `unemployment` data set gives us information about percentages of unemployment given a year and month. The range of years is from `r min(pull(unemployment_df, year))` to `r max(pull(unemployment_df, year))`.

The combined data set `FiveThirtyEight` has `r nrow(FiveThirtyEight_df)` observations and `r ncol(FiveThirtyEight_df)` variables. The range of years is from `r min(pull(FiveThirtyEight_df, year))` to `r max(pull(FiveThirtyEight_df, year))` and features important variables including the S&P stock market index at close, unemployment percentages, and whether the President was a Democrat or Republican at the time.

